### 2014-11-12 Studies/2011_San_Francisco/Report/San_Francisco_Fire.tex
San Francisco: Edits based on reviewer comments

% PDF metadata
\hypersetup{
  pdftitle={Simulation of a Fire in a Hillside Residential Structure - San Francisco, CA},
  pdfauthor={Overholt, Kristopher J., Weinschenk, Craig G., and Madrzykowski, Daniel},
  pdfsubject={The Fire Dynamics Simulator (FDS) fire model, which is developed and maintained by the National Institute of Standards and Technology (NIST), was used to provide insight into the dynamics of a fire that occurred on June 2, 2011, within a multi-level, single-family residential structure in San Francisco, CA, that resulted in the death of two firefighters. The inputs for the FDS simulation are documented in this report and are based on the fire scenario, including the building geometry, interior furnishings, and ventilation conditions. The fire started in the basement and resulted in ventilation limited (fuel rich) fire conditions in the basement area. The fire originated in the basement, and the interior stairwell acted as a chimney for hot gases in the basement to flow towards areas of lower pressure via the interior stairwell towards the front of the structure. After the rear basement windows began to fail, the combustion gases at elevated temperature and pressure in the basement flowed upwards towards lower pressure areas via the stairwell. The temperature of the gases in the interior stairwell was estimated to be in excess of 700~$^{\circ}$C (1300~$^{\circ}$F). Two firefighters were located in the flow path between the basement and the doors on the front side of the structure, were exposed to these elevated temperatures, and later died as a result of their injuries.},
  pdfkeywords={flow path; fire dynamics; LODD; fire modeling; FDS simulation}
}

@@

The fire originated in the basement, and the interior stairwell acted as a chimney for hot gases in the basement to flow towards areas of lower pressure via the interior stairwell towards the front of the structure. After the rear basement windows began to fail, the combustion gases at elevated temperature and pressure in the basement flowed upwards towards lower pressure areas via the stairwell. The temperature of the gases in the interior stairwell was estimated to be in excess of 700~$^{\circ}$C (1300~$^{\circ}$F). Two firefighters were located in the flow path between the basement and the doors on the front side of the structure, were exposed to these elevated temperatures, and later died as a result of their injuries.

@@

The account of events for this incident was documented in NIOSH Fire Fighter Fatality Investigation and Prevention Program Report \#F2011-13~\cite{NIOSH:Bowyer2}. For the purpose of this analysis, the details regarding the timeline are considered to be approximate values. The following narrative of the incident was extracted from the NIOSH incident report~\cite{NIOSH:Bowyer2}:
\begin{quote}
On June 02, 2011, a 48 year-old career lieutenant and a 53 year-old fire fighter/paramedic died in a multi-level residential structure fire while searching for the seat of the fire. Note: The residential structure where the fatalities occurred was built on a significantly sloped hillside common throughout the city. The fire floor was one floor below street level. Six companies and three command chiefs were dispatched to a report of an electrical fire at a residential home.

@@

Fire Dynamics Simulator~\cite{FDS_Users_Guide} is a computational fluid dynamics (CFD) model that solves a form of the Navier-Stokes equations appropriate for low-speed, thermally driven flow with an emphasis on smoke and heat transport from fires. Within a CFD model, the room or building is divided into small three-dimensional rectangular control volumes or computational cells. The cells are contained together within one larger volume known as the computational domain. The CFD model computes the density, velocity, temperature, pressure, and species concentration of the gas in each cell. Based on the laws of conservation of mass, momentum, and energy, the model tracks the generation and movement of fire gases. One of the most important advantages of FDS is that it is mathematically verified~\cite{FDS_Verification_Guide} and validated against fire test data~\cite{FDS_Validation_Guide} to ensure that it provides the expected results, given sufficient input data. A complete description of the FDS model is provided in the FDS Technical Reference Guide~\cite{FDS_Math_Guide}.

Smokeview is a software tool designed to visualize simulation results from FDS~\cite{Smokeview_Users_Guide}. Smokeview visualizes smoke and other attributes of the fire simulation using traditional scientific methods such as displaying tracer particle flow, two dimensional (2D) or three dimensional (3D) shaded contours of gas flow data such as temperature and flow vectors showing flow direction and magnitude. Smokeview has the ability to visualize fire and smoke by displaying a series of partially transparent planes where the transparencies in each plane (at each grid node) are determined from soot densities computed by FDS. Smokeview can also visualize data at particular snapshots in time using 2D or 3D contours of data that show temperature, flow direction, and flow magnitude.

Input data from various sources must be collected and documented to simulate a fire using FDS or any other fire model. For the simulation results presented in this document, information was obtained from two primary sources. The following information was gathered from the fire scene: the geometry of the building and the compartments being modeled, the size and location of exterior and interior ventilation openings, and documentation of fire damage to the building. The following information was gathered from witnesses, first responders, reports, and recorded media such as fire ground radios or photographs: information on the fire development timeline, the sequence and approximate timing of ventilation openings to the outside, and weather conditions.

@@

The analysis of the simulation results is focused on the interior conditions on the first floor where the downed firefighters were located and the conditions that resulted in the establishment of a flow path in the interior stairwell leading into the basement. Based on the goals of the analysis and the information collected, several assumptions were made. For example, the fire was in the early growth stage when firefighters arrived~\cite{NIOSH:Bowyer2}; however, the exact ignition time, fire growth rate, and heat release rate (HRR) of the fire were not known but were estimated using documentation of the potential fuel load within the structure.

@@

The structure was built on a sloped landscape such that, from the front side of the structure, two stories were located above the street level and two stories were located below the street level. The first and second floor (i.e., the two stories above street level) had a total interior floor area of 136~m$^2$ (1,468~ft$^2$). The interior walls were composed of gypsum board over insulated wood framing. The entrance stairs located on the front side of the structure led to the front door and an interior landing, which allowed access to the first floor (ten~stairs down) or the second floor (five~stairs up).

The basement (i.e., the first story below street level) had a total floor area of 121~m$^2$ (1301~ft$^2$) and consisted of a finished living area and a large utility area. A laundry area was located on a landing on the interior stairwell between the first floor and basement. The basement was accessible via this interior set of stairs or via an exterior door on the left side of the structure. The basement had an exterior balcony on the rear side of the structure. The sub-basement level (i.e., the second story below street level) was not connected to the other three levels of the house, contained no windows, and was only accessible via an exterior door of the left side of the structure. The exterior entrances to the basement and sub-basement levels were connected via exterior wooden stairs on the left side of the structure.

@@

The basement primarily contained upholstered polyurethane furniture, and the balcony was constructed of wood. The upholstered furniture items located in the middle of the basement living room were very large and flat, similar to the size and geometry of a mattress. Therefore, the HRR of the initial couch fire was based on an experimentally measured HRR for a mattress fire~\cite{madrzykowski2009fire} to represent the fire growth rate of a similar household furniture item. The HRRs of the additional furniture items were estimated based on previous experimental work that characterized the HRR of similar upholstered furniture fuel items~\cite{madrzykowski2009fire,Madrzykowski:2,Babrauskas:1}. Other combustible items and furnishings such as tables, stools, bookshelves, and carpet were present in the basement living room. However, to simplify the analysis and specification of the design fire, only the large upholstered furniture fuel items and balcony were included in the simulation because they were considered to have contributed the largest amount of energy in this fire scenario.

@@

Since the exact time at which the fire started is not known relative to the notification of a fire or how quickly the fire spread, several simplifying assumptions were made. One assumption is that the initial couch fire in the basement was specified to start at the beginning of the simulation. The duration of the fire simulation was 9~min (540~s). The beginning of the simulation (0~s) was selected to correspond to an on-scene time of 10:53:00, which was approximately five minutes after the fire department arrived on scene and approximately one minute before the E26 firefighting crew stated that the fire was located below the first floor. The end of the simulation (540~s) was selected to correspond to an on-scene time of 11:02:00, which was the time at which substantial fire suppression operations were initiated (see Table~\ref{tab:incident_timeline}).

@@

\item The secondary fires in the basement were divided among four additional furniture items (two couches, a lounge, and a chair). The HRRs of these additional furniture items were estimated based on previous experimental work that characterized the HRR of similar upholstered furniture fuel items, which reported HRRs of upholstered chairs and couches ranging from 2~MW to 5~MW~\cite{madrzykowski2009fire,Madrzykowski:2,Babrauskas:1,Janssens:2012}. The two large couches in the middle of the basement were specified with a peak HRR of 5~MW for each item. The lounge and chair located at the perimeter of the basement were specified with a peak HRR of 2~MW for each item. Therefore, the peak HRR of all four furniture items was a total of 14~MW.

@@

Polyurethane fuel was specified for the interior basement fuel items, which can be represented by the chemical formula, $\C\H_{1.8}\O_{0.30}\N_{0.05}$, with specified product yields of soot ($y_{\mathrm{s}}=0.131 \; \mathrm{kg}/\mathrm{kg}$) and carbon monoxide ($y_{\mathrm{CO}}=0.010 \; \mathrm{kg}/\mathrm{kg}$)~\cite{SFPE:Tewarson}. The product yields are expressed in terms of the amount soot or carbon monoxide emitted per unit mass of fuel consumed~(kg/kg) and can be found in the Society of Fire Protection Engineers (SFPE) Handbook~\cite{SFPE:Tewarson}. A balanced chemical reaction for polyurethane combustion can be written as:
\begin{multline}
\C\H_{1.8}\O_{0.30}\N_{0.05} + 5.22(0.208\,\O_{2} + 0.783\,\N_{2} + 8.34 \times 10^{-3}\,\H_{2}\O + 3.87 \times 10^{-4}\,\C\O_{2}) \\
\rightarrow 6.06(0.129\,\C\O_{2} + 0.156\,\H_{2}\O + 0.001\,\C\O + 0.035\,\C + 0.679\,\N_{2})
\label{eq:fuel_polyurethane}
\end{multline}

@@

Wood fuel was specified for the exterior basement balcony fuel items, which can be represented by the chemical formula, $\C\H_{1.7}\O_{0.74}\N_{0.002}$, with specified product yields of soot ($y_{\mathrm{s}}=0.015 \; \mathrm{kg}/\mathrm{kg}$) and carbon monoxide ($y_{\mathrm{CO}}=0.004 \; \mathrm{kg}/\mathrm{kg}$)~\cite{SFPE:Tewarson}. A balanced chemical reaction for wood combustion can be written as:
\begin{multline}
\C\H_{1.7}\O_{0.74}\N_{0.002} + 4.91(0.208\,\O_{2} + 0.783\,\N_{2} + 8.34 \times 10^{-3}\,\H_{2}\O + 3.87 \times 10^{-4}\,\C\O_{2}) \\
\rightarrow 5.74(0.168\,\C\O_{2} + 0.155\,\H_{2}\O + 6.37 \times 10^{-4}\,\C\O + 5.57 \times 10^{-3}\,\C + 0.670\,\N_{2})
\label{eq:fuel_wood}
\end{multline}
The heat of combustion of polyurethane and wood were specified as 26,200~kJ/kg and 16,400~kJ/kg, respectively, based on data provided in the SFPE Handbook~\cite{SFPE:Tewarson}. The heat of combustion represents the amount of energy released per unit mass of the fuel. The following text defines Eqs.~\ref{eq:fuel_polyurethane} and \ref{eq:fuel_wood} for an FDS input file with the two fuels and reaction properties discussed above:

@@

Note that using the input lines above will invoke the simple chemistry reaction mechanism in FDS in which fuel and air react to form only CO$_2$, CO, H$_2$O, soot, and N$_2$. If the inclusion of other combustion products is desired, then the user must explicitly define those species and the chemical reaction that produces them~\cite{FDS_Users_Guide}. Based on the above input lines, FDS uses the default, mixing-controlled fast chemistry combustion model. This mechanism states that the rate of fuel consumption is proportional to both the local limiting reactant concentration and the local rate of mixing, and extinction is based on the critical flame temperature approach~\cite{FDS_Math_Guide}. While FDS provides users with the option to use a more complex finite-rate combustion mechanism, there is not sufficient evidence in this case to justify deviating from the default specifications.

Another simplification of the model is that the sources of the gas phase polyurethane and wood fuel have constant areas at fixed locations (five furniture items in the basement interior for the polyurethane fuel, and the exterior basement balcony and siding for the wood fuel).

A time-dependent burning rate was specified for each of the fuel packages in terms of a fuel mass flux. The fuel mass flux is the amount of fuel vapor per unit area that is released from the surface of each fuel package shown in Fig.~\ref{fig:fuel_placement}. The specified time-dependent fuel mass flux for each fuel package corresponds to the HRRs that were described earlier. The HRR from each individual fuel package contributes to the total HRR shown in Fig.~\ref{fig:hrr}. FDS then determines the amount of combustion that occurs throughout the simulation based on the amount of available fuel and oxygen at a given location. Each fuel item was modeled as a 3D object with surfaces from which a specified amount of fuel flows into the domain, mixes with air, and burns.

@@

While the fires were represented as fuel sources with a constant area (Section~\ref{sec:fires}), from a heat transfer perspective, it is important to define the material properties of the ceiling, walls, and floors (density, thermal conductivity, specific heat, and thickness) to account for heat transfer and energy storage. In this study, the material properties of gypsum board~\cite{WAKILI2007} were specified on the finished walls and ceilings of the structure. Table~\ref{tab:model_parameters} shows the material properties of the gypsum board that were used in the simulation.

@@

The simulation in this study accounted for changes in ventilation due to a combination of fire department operations (opening doors) and fire acting on the structure (breaking windows). The ventilation times and ventilation areas represent our best understanding of the incident. The times of ventilation changes for doors and windows are provided in Table~\ref{tab:ventilation_timeline}.

@@

Structure leakage has been shown to be important when modeling enclosure fires~\cite{beal2009}. The open front door at the start of the simulation and the vents created during the simulation are significantly larger than the total effective area of leakage of a structure of this type. Therefore, any leakage in the structure would have a negligible impact on the fluid mechanics within the structure and was not included in this study.

@@

As discussed in Section~\ref{sec:geometry}, the structure had exterior dimensions of 8.5~m (28~ft) by 17.7~m (58~ft) and a height of 5.9~m (19~ft) above street level. To ensure adequately resolved fluid flow in and out of the structure, the computational domain was extended beyond the volume of the structure. The dimensions of the computational domain were 11.5~m (38~ft) by 22~m (72~ft) and a height of 13.6~m (45~ft). A grid resolution of 10~cm (3.9~in) was used, which resulted in a total of approximately 3.4 million computational cells. As a result, all of the obstructions, ventilation openings (holes, doors, and windows), and fire areas snapped to the nearest 10~cm. Figure~\ref{fig:SMV_exterior_grid} shows the front and rear sides of the structure rendered in Smokeview with the 10~cm grid cells.

@@

\begin{table}[!ht]
\caption[Relevant fire model input parameters]
{Relevant fire model input parameters}
\begin{tabular}{lll}
\toprule
Parameter                                    &  Description                                  &  Discussion                        \\
\midrule
Simulation Time                              &  9~min                                        &  --                                \\
Grid Cell Size                               &  10~cm                                        &  Section~\ref{sec:numerical_mesh}  \\
Ambient Temperature$^*$                      &  20~$^{\circ}$C (68~$^{\circ}$F)              &  --                                \\
Reaction: Polyurethane~\cite{SFPE:Tewarson}  &  Formula: $\C\H_{1.8}\O_{0.30}\N_{0.05}$      &  Section~\ref{sec:fires}           \\
                                             &  Soot Yield: 0.131~$\mathrm{kg}/\mathrm{kg}$  &                                    \\
                                             &  CO Yield: 0.010~$\mathrm{kg}/\mathrm{kg}$    &                                    \\
                                             &  Heat of combustion: 26,200~kJ/kg             &                                    \\
Reaction: Wood~\cite{SFPE:Tewarson}          &  Formula: $\C\H_{1.7}\O_{0.74}\N_{0.002}$     &  Section~\ref{sec:fires}           \\
                                             &  Soot Yield: 0.015~$\mathrm{kg}/\mathrm{kg}$  &                                    \\
                                             &  CO Yield: 0.004~$\mathrm{kg}/\mathrm{kg}$    &                                    \\
                                             &  Heat of combustion: 16,400~kJ/kg             &                                    \\
Peak HRRs                                    &  Initial couch fire: 5~MW                     &  Section~\ref{sec:fires}           \\
                                             &  Basement flashover fire: 14~MW               &                                    \\
                                             &  Basement balcony fire: 1~MW                  &                                    \\
Material: Gypsum Board~\cite{WAKILI2007}     &  Thermal conductivity: 0.28~\si{W/(m.K)}      &  Section~\ref{sec:materials}       \\
                                             &  Density: 810~kg/m$^3$                        &                                    \\
                                             &  Specific heat: 1.0~\si{kJ/(kg.K)}            &                                    \\
                                             &  Thickness: 1.2~\si{cm}                       &                                    \\
\bottomrule
\end{tabular}
\footnotesize
\\ $^{*}$Initial interior temperature was assumed to be 20~$^{\circ}$C; actual exterior temperature was 14~$^{\circ}$C~\cite{NIOSH:Bowyer2}.
\normalsize
\label{tab:model_parameters}
\end{table}

@@

In an FDS simulation, a HRR is specified, which results in a specified amount of fuel vapor (or pyrolyzate) being released from a fuel surface. FDS then determines the amount of combustion that occurs throughout the simulation based on the amount of available fuel and oxygen at a given location. As a result, the prescribed HRR that was input into the simulation can be different than the HRR predicted by the simulation based on the ventilation conditions.

Figure~\ref{fig:hrr_w_fds} shows the HRR vs. time based on the fires that were described in Section~\ref{sec:fires}. In this figure, the solid line represents the specified (or prescribed) HRR that was input into the simulation, and the dashed line represents the predicted HRR that was output from the simulation based on the ventilation conditions. The vertical dashed line represents the time at which the first basement window on the rear side of the structure failed (342~s). Additional windows on the rear side of the basement continued to fail within the next two minutes.

The maximum size of the fire used in this study was selected such that the conditions would become ventilation limited to minimize computational time while aligning with observations from the actual fire scenario. Note that the rear window failures were manually specified in accordance with the event timeline rather than specifying failure temperature criteria.

>>>>

% PDF metadata
\hypersetup{
  pdftitle={Simulation of a Fire in a Hillside Residential Structure - San Francisco, CA},
  pdfauthor={Overholt, Kristopher J., Weinschenk, Craig G., and Madrzykowski, Daniel},
  pdfsubject={The Fire Dynamics Simulator (FDS) fire model, which is developed and maintained by the National Institute of Standards and Technology (NIST), was used to provide insight into the dynamics of a fire that occurred on June 2, 2011, within a multi-level, single-family residential structure in San Francisco, CA, that resulted in the death of two firefighters. The inputs for the FDS simulation are documented in this report and are based on the fire scenario, including the building geometry, interior furnishings, and ventilation conditions. The fire started in the basement and resulted in ventilation limited (fuel rich) fire conditions in the basement area. The fire originated in the basement, and the interior stairwell acted as a chimney for hot gases in the basement to flow towards regions of lower pressure via the interior stairwell towards the front of the structure. After the rear basement windows began to fail, the combustion gases at elevated temperature and pressure in the basement flowed upwards towards lower pressure regions via the stairwell. The temperature of the gases in the interior stairwell was estimated to be in excess of 700~$^{\circ}$C (1300~$^{\circ}$F). Two firefighters were located in the flow path between the basement and the doors on the front side of the structure, were exposed to these elevated temperatures, and later died as a result of their injuries.},
  pdfkeywords={flow path; fire dynamics; LODD; fire modeling; FDS simulation}
}

@@

The fire originated in the basement, and the interior stairwell acted as a chimney for hot gases in the basement to flow towards regions of lower pressure via the interior stairwell towards the front of the structure. After the rear basement windows began to fail, the combustion gases at elevated temperature and pressure in the basement flowed upwards towards lower pressure regions via the stairwell. The temperature of the gases in the interior stairwell was estimated to be in excess of 700~$^{\circ}$C (1300~$^{\circ}$F). Two firefighters were located in the flow path between the basement and the doors on the front side of the structure, were exposed to these elevated temperatures, and later died as a result of their injuries.

@@

On the morning of June 02, 2011, the San Francisco Fire Department was dispatched to a residence based on a report from the occupant of curtains on fire. The account of events for this incident was documented in NIOSH Fire Fighter Fatality Investigation and Prevention Program Report \#F2011-13~\cite{NIOSH:Bowyer2}. For this analysis, the details regarding the timeline are considered to be approximate values. The following narrative of the incident was extracted from the NIOSH incident report~\cite{NIOSH:Bowyer2}:
\begin{quote}
On June 02, 2011, a 48 year-old career lieutenant and a 53 year-old fire fighter/paramedic died in a multi-level residential structure fire while searching for the seat of the fire. Note: The residential structure where the fatalities occurred was built on a significantly sloped hillside common throughout the city. The fire floor was one floor below street level. Six companies and three command chiefs were dispatched to a report of an electrical fire at a residential home.

@@

Fire Dynamics Simulator~\cite{FDS_Users_Guide} is a computational fluid dynamics (CFD) model developed and maintained by NIST that solves a form of the Navier-Stokes equations appropriate for low-speed, thermally driven flow with an emphasis on smoke and heat transport from fires. Within a CFD model, the room or building is divided into small three-dimensional rectangular control volumes or computational cells. The cells are contained together within a larger volume known as the computational domain. The CFD model computes the density, velocity, temperature, pressure, and species concentration of the gas in each cell. Based on the laws of conservation of mass, momentum, and energy, the model tracks the generation and movement of fire gases. One of the most important aspects of FDS is that it is mathematically verified~\cite{FDS_Verification_Guide} and validated against fire test data~\cite{FDS_Validation_Guide} to ensure that it is accurate and provides the expected results, given appropriate input data. A complete description of the FDS model is provided in the FDS Technical Reference Guide~\cite{FDS_Math_Guide}.

Smokeview is a software tool designed to visualize simulation results from FDS~\cite{Smokeview_Users_Guide}. Smokeview visualizes smoke and other attributes of the fire simulation using advanced scientific methods such as displaying tracer particle flow and two dimensional (2D) or three dimensional (3D) shaded contours of gas flow data, including temperature and flow vectors showing flow direction and magnitude. Smokeview has the ability to visualize fire and smoke by displaying a series of partially transparent planes where the transparencies in each plane (at each grid node) are determined from soot densities computed by FDS. Smokeview can also visualize data at particular snapshots in time using 2D or 3D contours of data that show temperature, flow direction, and flow magnitude.

Input data from various sources must be collected and documented to simulate a fire using FDS or any other fire model. For the simulation results presented in this document, information was obtained from two primary sources. The following information was gathered from the fire scene: the geometry of the building and the compartments being modeled, the size and location of exterior and interior ventilation openings, and documentation of fire damage to the building. The following information was gathered from witnesses, first responders, reports from the San Francisco Fire Department and NIOSH, and recorded media such as fire ground radios or photographs: information on the fire development timeline, the sequence and approximate timing of ventilation openings to the outside, and weather conditions.

@@

The analysis of the simulation results is focused on the interior conditions on the first floor where the downed firefighters were located and the conditions that resulted in the establishment of a flow path in the interior stairwell leading into the basement. Based on the goals of the analysis and the information collected, several assumptions were made. For example, the fire was in the early growth stage when firefighters arrived~\cite{NIOSH:Bowyer2}; however, the exact ignition time, fire growth rate, and transient heat release rate (HRR) of the fire were not known. The potential fuel load within the structure was estimated from post-incident photographs and the timeline from the NIOSH incident report.

@@

The structure was built on a sloped landscape such that, from the front side of the structure, two stories were located at and above street level, and two stories were located below street level. The first floor and second floor (i.e., the two stories at and above street level) each had an interior floor area of 136~m$^2$ (1,470~ft$^2$). The interior walls were composed of gypsum board over insulated wood framing. The entrance stairs located on the front side of the structure led to the front door and an interior landing, which allowed access to the first floor (ten~stairs down) or the second floor (five~stairs up).

The basement (i.e., the first story below street level) had a total floor area of 121~m$^2$ (1300~ft$^2$) and consisted of a finished living area and a large utility area. A laundry area was located on a landing on the interior stairwell between the first floor and basement. The basement was accessible via this interior set of stairs or via an exterior door on the left side of the structure. The basement had an exterior balcony on the rear side of the structure. The sub-basement level (i.e., the second story below street level) was not connected to the other three levels of the house, contained no windows, and was only accessible via an exterior door of the left side of the structure. The exterior entrances to the basement and sub-basement levels were connected via exterior wooden stairs on the left side of the structure.

@@

The basement contained upholstered furniture composed primarily of polyurethane foam, and the balcony was constructed of wood. The upholstered furniture items located in the middle of the basement living room were very large and flat, similar to the size and geometry of a mattress. Therefore, the HRR of the initial couch fire was based on an experimentally measured HRR for a mattress fire~\cite{madrzykowski2009fire} to represent the fire growth rate of a similar household furniture item. The HRRs of the additional furniture items were estimated based on previous experimental work that characterized the HRR of geometrically similar upholstered furniture fuel items~\cite{madrzykowski2009fire,Madrzykowski:2,Babrauskas:1}. Other combustible items and furnishings such as tables, stools, bookshelves, and carpet were present in the basement living room. However, to simplify the analysis and specification of the design fire, only the large upholstered furniture fuel items and balcony were included in the simulation because they were considered to have contributed the largest amount of energy in this fire scenario.

@@

Since the exact time at which the fire started was not known relative to the notification of a fire or how quickly the fire spread, several simplifying assumptions were made. One assumption was that the initial couch fire in the basement was specified to start at the beginning of the simulation. The duration of the fire simulation was 9~min (540~s). The beginning of the simulation (0~s) was selected to correspond to an on-scene time of 10:53:00, which was approximately five minutes after the fire department arrived on scene and approximately one minute before the E26 firefighting crew stated that the fire was located below the first floor. The end of the simulation (540~s) was selected to correspond to an on-scene time of 11:02:00, which was the time at which substantial fire suppression operations were initiated (see Table~\ref{tab:incident_timeline}).

@@

\item The secondary fires in the basement were divided among four additional upholstered furniture items (two couches, a lounge chair, and a chair). The HRRs of these additional furniture items were estimated based on previous experimental work that characterized the HRR of geometrically similar upholstered furniture fuel items, which reported HRRs of upholstered chairs and couches ranging from 2~MW to 5~MW~\cite{madrzykowski2009fire,Madrzykowski:2,Babrauskas:1,Janssens:2012}. The two large couches in the middle of the basement were specified with a peak HRR of 5~MW for each item. The lounge and chair located at the perimeter of the basement were specified with a peak HRR of 2~MW for each item. Therefore, the peak HRR of all four furniture items was a total of 14~MW.

@@

Polyurethane foam was specified for the interior basement fuel items, which can be represented by the chemical formula, $\C\H_{1.8}\O_{0.30}\N_{0.05}$, with specified product yields of soot ($y_{\mathrm{s}}=0.131 \; \mathrm{kg}/\mathrm{kg}$) and carbon monoxide ($y_{\mathrm{CO}}=0.010 \; \mathrm{kg}/\mathrm{kg}$)~\cite{SFPE:Tewarson}. The product yields are constant and are expressed in terms of the amount soot or carbon monoxide emitted per unit mass of fuel consumed~(kg/kg) and can be found in the Society of Fire Protection Engineers (SFPE) Handbook~\cite{SFPE:Tewarson}. A balanced chemical reaction for the combustion of polyurethane foam in ambient air at 40~\% relative humidity can be written as:
\begin{multline}
\C\H_{1.8}\O_{0.30}\N_{0.05} + 5.22(0.208\,\O_{2} + 0.783\,\N_{2} + 8.34 \times 10^{-3}\,\H_{2}\O + 3.87 \times 10^{-4}\,\C\O_{2}) \\
\rightarrow 6.06(0.129\,\C\O_{2} + 0.156\,\H_{2}\O + 0.001\,\C\O + 0.035\,\C + 0.679\,\N_{2})
\label{eq:fuel_polyurethane}
\end{multline}

@@

Wood was specified for the exterior basement balcony fuel items, which can be represented by the chemical formula, $\C\H_{1.7}\O_{0.74}\N_{0.002}$, with specified product yields of soot ($y_{\mathrm{s}}=0.015 \; \mathrm{kg}/\mathrm{kg}$) and carbon monoxide ($y_{\mathrm{CO}}=0.004 \; \mathrm{kg}/\mathrm{kg}$)~\cite{SFPE:Tewarson}. A balanced chemical reaction for wood combustion can be written as:
\begin{multline}
\C\H_{1.7}\O_{0.74}\N_{0.002} + 4.91(0.208\,\O_{2} + 0.783\,\N_{2} + 8.34 \times 10^{-3}\,\H_{2}\O + 3.87 \times 10^{-4}\,\C\O_{2}) \\
\rightarrow 5.74(0.168\,\C\O_{2} + 0.155\,\H_{2}\O + 6.37 \times 10^{-4}\,\C\O + 5.57 \times 10^{-3}\,\C + 0.670\,\N_{2})
\label{eq:fuel_wood}
\end{multline}
The heat of combustion of polyurethane foam and wood were specified as 26,200~kJ/kg and 16,400~kJ/kg, respectively, based on data provided in the SFPE Handbook~\cite{SFPE:Tewarson}. The heat of combustion represents the amount of energy released per unit mass of the fuel. The following text defines Eqs.~\ref{eq:fuel_polyurethane} and \ref{eq:fuel_wood} for an FDS input file with the two fuels and reaction properties discussed above:

@@

Note that using the input lines above will invoke the simple chemistry reaction mechanism in FDS in which fuel and air react to form only CO$_2$, CO, H$_2$O, soot, and N$_2$. If the inclusion of other combustion products is desired, then the user must explicitly define those species and the chemical reaction that produces them~\cite{FDS_Users_Guide}. Based on the above input lines, FDS uses the default, mixing-controlled fast chemistry combustion model. This mechanism states that the rate of fuel consumption is proportional to both the local limiting reactant concentration and the local rate of mixing, and extinction is based on a critical flame temperature~\cite{FDS_Math_Guide}. While FDS provides users with the option to use a more complex finite-rate combustion mechanism, there is not a sufficient amount of information to justify deviating from the default specifications.

Another simplification of the model is that the sources of the gas phase polyurethane foam and wood fuel had constant areas at fixed locations (five furniture items in the basement interior for the polyurethane foam, and the exterior basement balcony and siding for the wood fuel).

A time-dependent burning rate was specified for each of the fuel packages in terms of a fuel mass flux. The fuel mass flux is the amount of fuel vapor per unit area that is released from the surface of each fuel package shown in Fig.~\ref{fig:fuel_placement}. The specified time-dependent fuel mass flux for each fuel package corresponds to the HRRs that were described earlier. The HRR from each individual fuel package contributed to the total HRR shown in Fig.~\ref{fig:hrr}. FDS then determined the amount of combustion that occurred throughout the simulation based on the amount of available fuel and oxygen at a given location. Each fuel item was modeled as a 3D object with surfaces from which a specified amount of fuel flowed into the domain, mixed with air, and burned.

@@

Whereas the fires were represented as fuel sources with a constant area (Section~\ref{sec:fires}), from a heat transfer perspective, it is important to define the material properties of the ceiling, walls, and floors (density, thermal conductivity, specific heat, and thickness) to account for heat transfer and energy storage. In this study, the material properties of gypsum board~\cite{WAKILI2007} were specified on the finished walls and ceilings of the structure. Table~\ref{tab:model_parameters} shows the material properties of the gypsum board that were used in the simulation.

@@

The simulation in this study accounted for changes in ventilation due to a combination of fire department operations (opening doors) and fire acting on the structure (breaking windows). The ventilation times and ventilation areas represent our best understanding of the incident. The estimated times of ventilation changes for doors and windows are provided in Table~\ref{tab:ventilation_timeline}.

@@

Air leakage in structures has been shown to be an important consideration when modeling enclosure fires~\cite{beal2009}. The open front door at the start of the simulation and the vents created during the simulation are significantly larger than the total effective area of leakage of a structure of this type. Therefore, any leakage in the structure would have a relatively negligible impact on the fluid mechanics and combustion within the structure and was not included in this study.

@@

As discussed in Section~\ref{sec:geometry}, the structure had exterior dimensions of 8.5~m (28~ft) by 17.7~m (58~ft) and a height of 5.9~m (19~ft) above street level. To ensure adequately resolved fluid flow in and out of the structure, the computational domain was extended beyond the volume of the structure. The dimensions of the computational domain were 11.5~m (38~ft) by 22~m (72~ft) and a height of 13.6~m (45~ft). A grid resolution of 10~cm (3.9~in) was used, which resulted in a total of approximately 3.4 million computational cells. As a result, all of the obstructions, ventilation openings (doors and windows), and fire areas were snapped by FDS to the nearest 10~cm. Figure~\ref{fig:SMV_exterior_grid} shows the front and rear sides of the structure rendered in Smokeview with the 10~cm grid cells.

@@

\begin{table}[!ht]
\caption[Fire model input parameters]
{Fire model input parameters}
\begin{tabular}{lll}
\toprule
Parameter                                         &  Description                                  &  Discussion                        \\
\midrule
Simulation Time                                   &  9~min                                        &  --                                \\
Grid Cell Size                                    &  10~cm                                        &  Section~\ref{sec:numerical_mesh}  \\
Ambient Temperature$^*$                           &  20~$^{\circ}$C (68~$^{\circ}$F)              &  --                                \\
Reaction: Polyurethane Foam~\cite{SFPE:Tewarson}  &  Formula: $\C\H_{1.8}\O_{0.30}\N_{0.05}$      &  Section~\ref{sec:fires}           \\
                                                  &  Soot Yield: 0.131~$\mathrm{kg}/\mathrm{kg}$  &                                    \\
                                                  &  CO Yield: 0.010~$\mathrm{kg}/\mathrm{kg}$    &                                    \\
                                                  &  Heat of combustion: 26,200~kJ/kg             &                                    \\
Reaction: Wood~\cite{SFPE:Tewarson}               &  Formula: $\C\H_{1.7}\O_{0.74}\N_{0.002}$     &  Section~\ref{sec:fires}           \\
                                                  &  Soot Yield: 0.015~$\mathrm{kg}/\mathrm{kg}$  &                                    \\
                                                  &  CO Yield: 0.004~$\mathrm{kg}/\mathrm{kg}$    &                                    \\
                                                  &  Heat of combustion: 16,400~kJ/kg             &                                    \\
Peak HRRs                                         &  Initial couch fire: 5~MW                     &  Section~\ref{sec:fires}           \\
                                                  &  Basement flashover fire: 14~MW               &                                    \\
                                                  &  Basement balcony fire: 1~MW                  &                                    \\
Material: Gypsum Board~\cite{WAKILI2007}          &  Thermal conductivity: 0.28~\si{W/(m.K)}      &  Section~\ref{sec:materials}       \\
                                                  &  Density: 810~kg/m$^3$                        &                                    \\
                                                  &  Specific heat: 1.0~\si{kJ/(kg.K)}            &                                    \\
                                                  &  Thickness: 1.2~\si{cm}                       &                                    \\
\bottomrule
\end{tabular}
\footnotesize
\\ $^{*}$Initial interior temperature was assumed to be 20~$^{\circ}$C; actual exterior temperature was 14~$^{\circ}$C~\cite{NIOSH:Bowyer2}.
\normalsize
\label{tab:model_parameters}
\end{table}

@@

In FDS, a HRR is specified, which results in a specified amount of fuel vapor (or pyrolyzate) being released from a fuel surface. FDS then determines the amount of combustion that occurs throughout the simulation based on the amount of available fuel and oxygen at a given location. As a result, the prescribed HRR that was input into the simulation can be different than the HRR calculated by FDS based on the ventilation conditions.

Figure~\ref{fig:hrr_w_fds} shows the HRR vs. time based on the fires that were described in Section~\ref{sec:fires}. In this figure, the solid line represents the prescribed HRR that was input into the simulation (based on a prescribed mass flux from the fuel items), and the dashed line represents the HRR that was calculated by FDS based on the ventilation conditions. The vertical dashed line represents the time at which the first basement window on the rear side of the structure failed (342~s). Additional windows on the rear side of the basement continued to fail within the next two minutes.

The maximum size of the fire specified in this study was selected such that the conditions aligned with observations from the actual fire incident. Note that the rear window failures were manually specified in accordance with the event timeline rather than specifying failure temperature criteria.
